Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-02-03 manuscript/manuscript.Rmd
Pete round 2 revisions

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while maintaining diversity that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building corals plays a major role in their biology and ecology, as the identity and functional performance of individual symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts within a coral may have important functional consequences [@Loram:2007p3688; @Cunning:2015ja]. Moreover, plasticity and flexibility of these communities within species and individuals over time is an important trait modulating the sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond adaptively by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding the ecological performance of corals depends on accurately and comprehensively describing their *Symbiodinium* communities.

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2. Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g. [@Dover:1986ju]). In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level. Moreover, due to the high volume of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance our understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations. Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity may underestimate diversity by collapsing species known to be different. True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species. The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

@@

  A total of 84 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St. John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave exposure (J m^-3^). These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition [@Houlbreque:2009gh]. SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite for each month of the year, averaged over the period 2003-2010. We used as input the monthly climatologies available at the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates, values from the nearest available pixel were used. Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig. 1).

  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al. 2012) for 32 fetch directions and coastline data at full resolution. Mean wave exposure at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

@@

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "Symbiodinium".

@@

  Beta diversity (sensu [@Anderson:2006fs]) was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

@@

  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The sequence data generated from these samples comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1. From 2003 to 2010, the north shore of St. John was characterized by slightly lower sea surface temperatures, higher chlorophyll a concentrations, and lower wave exposure, relative to the south shore.

@@

# PLOT MAP
par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
plot(stjmap, lwd=1.5, col="black",
     xlim=c(-64.77, -64.70), ylim=c(18.303, 18.370))
#text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.78, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.78, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black", cex=0.8)
  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
})
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)

@@

#Wind exposure
we <- list(N=3.234787362,	S=4.025022708)
plot(NA, xlim=c(3,4.2), ylim=c(0.5,2.5), xlab="Wave exposure (J m-3)", ylab="", bty="n", yaxt="n")
points(c(we$N, we$S), c(2,1), pch=19)
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)

@@

```{r Table1}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

#### Symbiodinium community composition

@@

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs. south). Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
```

@@

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

  Patterns of association between hosts and *Symbiodinium* were analyzed by constructing three networks focusing on "abundant" (>1% relative abundance; Fig. 4), "dominant" (>50% relative abundance; Fig. 5A), and "background" (<1% relative abundance; Fig. 5B) symbionts in each host species, based on 97% within-sample clustering. (Networks for individuals within each species are presented in Supplementary Materials). In the abundant symbiont network (Fig. 4), `r sum(V(abunet)$type==2)` OTUs were observed. *D. strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F. fragum* (n=`r abu.Nedges["fragum"]`), *M. alcicornis* (n=`r abu.Nedges["alcicornis"]`), and *D. cylindrus* (n=`r abu.Nedges["cylindrus"]`); these primarily comprised OTUs closely related to B1 and B19. In contrast, only a single OTU occurred at >1% relative abundance in both *P. astreoides* (A4) and *M. cavernosa* (C3). All coral species hosted at least one "abundant" *Symbiodinium* OTU that was also abundant in at least one other coral species, except for *S. radians*. This species was dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.

>>>>

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-tolerance of their coral hosts, making the description of *Symbiodinium* communities in corals (and metacommunities on reefs) a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented scale, yet high intragenomic variability at this locus complicates the resolution of biologically meaningful diversity. Using samples of *Symbiodinium* from ten host taxa on shallow reefs on the north and south shores of St. John, US Virgin Islands, we demonstrate that generating operational taxonomic units by clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while preserving diversity that is more likely interspecific. We utilized this ‘within-sample clustering’ approach to create metacommunity networks of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages occupying 'dominant' and 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations with *Symbiodinium*, although no differences in symbiont communities were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  The composition of symbiotic algal communities (*Symbiodinium* spp.) in reef-building scleractinian and milleporine corals plays a major role in their biology and ecology, as the identity and functional performance of the symbionts influences emergent properties of the holobiont, including its photobiology, energetics, growth rates, and susceptibility to stress [@Rowan:2004p3383; @Cantin:2009p4376; @Jones:2011dt]. Even slight differences in the relative abundance of different symbionts may have important functional consequences for the host [@Loram:2007p3688; @Cunning:2015ja]. Moreover, variation in these communities among individuals within a host species, and within individuals over time, is an important trait modulating sensitivity of corals to environmental stress [@Putnam:2012wv] and their ability to respond in beneficial ways to changing environmental conditions by 'shuffling' or 'switching' symbionts [@Baker:2003p200]. Therefore, understanding the ecological performance of corals will greatly benefit from accurate and comprehensive descriptions of their *Symbiodinium* communities.

  *Symbiodinium* identity is primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; @Santos:2003kt; @Pochon:2012we]. Together, these markers have been used to identify nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which have been further divided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, ecological surveys of *Symbiodinium* diversity must still rely on commonly-used marker genes, such as ITS2. Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize *Symbiodinium* communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses may create datasets consisting of millions of sequence reads and hundreds of thousands of distinct sequence variants [@Ziegler:2017cs], which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within the genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly-repeating ribosomal DNA array creates multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g. [@Dover:1986ju]). In fact, concerted evolution may mask species divergence within *Symbiodinium* by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker for *Symbiodinium*. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level. Moreover, due to the large quantity of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2, it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys of *Symbiodinium*.

  Here, we describe a *Symbiodinium* metacommunity associated with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' (i.e. clustering sequences at 97% similarity within each sample independently) to generate biologically relevant operational taxonomic units (OTUs) from ITS2 metabarcoding data. Specifically, this approach addresses the fact that dominant ITS2 sequences from different *Symbiodinium* species may be more similar to one another than intragenomic variants within one *Symbiodinium* [@Thornhill:2013dx; @Arif:2014gk; @Parkinson:2015gt]. Therefore, clustering all sequences together underestimates diversity by collapsing different species into the same OTU, while treating each ITS2 sequence as a unique *Symbiodinium* type overestimates diversity due to intragenomic variation. Within-sample clustering may better approximate true diversity by exploiting key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to be intragenomic variants, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using within-sample clustering, we analyzed *Symbiodinium* communities in ten host species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species. The large quantity of data, and the kinds of analytical approaches facilitated by ITS2 metabarcoding, have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, the present study demonstrates how ecological knowledge can inform bioinformatic analyses using marker genes for which sequence diversity does not map directly to species.

@@

  Eighty-four tissue samples were collected from scleractinian and milleporine corals sampled at seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies to be sampled were selected haphazardly in upward-facing locations exposed to ambient irradiance. Tissue biopsies (~5-mm diameter and ~5-mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St. John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and integrated wave exposure on a power scale (J m^-3^). These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and conditions known to affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and mediate the supply of food resources that play important roles in coral nutrition [@Houlbreque:2009gh]. SST and chlorophyll were determined using data from the MODIS sensor on the Aqua satellite. SST was evaluated from nighttime records, and chlorophyll from ocean colour, both of which were obtained at 1 km spatial resolution for each month, averaged over the years 2003-2010, from the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates (above), values from the nearest available pixel were used. Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig. 1).

  Wind-driven wave exposure on a power scale for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch. To calculate wave exposure, wind speed and direction at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al. 2012) for 32 fetch directions (equally distributed through 360°) and coastline data at full resolution (xx km). Total wave exposure (summed over all directions) was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

@@

  Paired reads from each sample (provided in .fastq format by Research and Testing Laboratory) were merged using illumina-utils software [@Eren:2013hh] with an enforced Q30-check and an overlap ≥150 bp with ≤3 mismatches required to retain a sequence. Chimeric sequences were removed using usearch61 [@Edgar:2010cv] implemented in QIIME [@Caporaso:2010jf]. Primers were trimmed using cutadapt [@Martin:2011eu] allowing 3 mismatches, and only sequences with both forward and reverse primer matches and length ≥250 bp after trimming were retained. Subsequently, three different clustering approaches, each based on the uclust algorithm [@Edgar:2010cv] and implemented in QIIME, were used to group ITS2 sequences into operational taxonomic units (OTUs): 1) clustering at 100% identity, 2) clustering at 97% identity across samples (i.e., sequences from all samples clustered together), and 3) clustering at 97% identity within samples (i.e., sequences from each sample clustered independently). For each 97% cluster, the most abundant sequence variant was chosen as the representative sequence, and within-sample clusters were merged across samples if their representative sequences were 100% identical. After removing singleton clusters, representative sequences for each OTU were assigned taxonomy by searching a custom reference database of *Symbiodinium* ITS2 sequences using the Needleman-Wunsch global alignment algorithm implemented in the Biostrings package [@BiostringsStringo:Fbauxymn] in R [@RALanguageandEn:2014wf]. Each OTU was then assigned a name corresponding to the reference sequence(s) with the highest alignment score; if the match was <100%, the assignment was given a unique superscript. If the match was <90%, the sequence was blasted to NCBI and omitted if the top hit did not contain "*Symbiodinium*".

@@

  Beta diversity (sensu [@Anderson:2006fs]) was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using the betadisper function in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

@@

  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The sequence data generated from these samples comprised 1,490,813 sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

  The environmental conditions broadly characterizing the north and south shores are presented in Fig. 1. From 2003 to 2010, the north shore of St. John was characterized by slightly lower sea surface temperatures, higher chlorophyll a concentrations, and lower wave exposure, relative to the south shore.

@@

# PLOT MAP
par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
plot(stjmap, lwd=1.5, col="gray40",
     xlim=c(-64.77, -64.70), ylim=c(18.303, 18.370))
#text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.78, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.78, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black", cex=0.8)
  text(lon, lat, name, pos=c(4,2,4,4,4,1,3), cex=0.8)
})
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)

@@

#Wave exposure
we <- list(N=3.234787362,	S=4.025022708)
barplot(c(we$N, we$S), horiz=T, xlim=c(0,5), ylab="", xlab="Wave exposure (ln J m-3)")
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1, xpd=NA)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5, xpd=NA)

@@

```{r Table1, cache=F}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

#### *Symbiodinium* community composition

@@

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs. south). Individual segments of each bar represent 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
#png(file="crap.png", width = 4, height=6, units="in", res=300)
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=1)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
#dev.off()
```

@@

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
shorestats97bs$Species <- str_replace_all(
  shorestats97bs$Species,
  c("alcicornis"="*Millepora alcicornis*",
    "annularis"="*Orbicella annularis*",
    "astreoides"="*Porites astreoides*",
    "cavernosa"="*Montastraea cavernosa*",
    "cylindrus"="*Dendrogyra cylindrus*",
    "fragum"="*Favia fragum*",
    "furcata"="*Porites furcata*",
    "radians"="*Siderastrea radians*",
    "siderea"="*Siderastrea siderea*",
    "strigosa"="*Pseudodiplora strigosa*")
)
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

  Patterns of association between hosts and *Symbiodinium* were analyzed by constructing three networks focusing on "abundant" (>1% relative abundance; Fig. 4), "dominant" (>50% relative abundance; Fig. 5A), and "background" (<1% relative abundance; Fig. 5B) symbionts in each host species, based on 97% within-sample clustering. (Networks for individuals within each species are presented in Supplementary Materials). In the network for abundant symbionts (Fig. 4), `r sum(V(abunet)$type==2)` OTUs were observed. *P. strigosa* associated with the greatest number of OTUs (n=`r abu.Nedges["strigosa"]`), followed by *F. fragum* (n=`r abu.Nedges["fragum"]`), *M. alcicornis* (n=`r abu.Nedges["alcicornis"]`), and *D. cylindrus* (n=`r abu.Nedges["cylindrus"]`); these primarily comprised OTUs closely related to B1 and B19. In contrast, only a single OTU occurred at >1% relative abundance in both *P. astreoides* (A4) and *M. cavernosa* (C3). All coral species hosted at least one "abundant" *Symbiodinium* OTU that was also abundant in at least one other coral species, except for *S. radians*. This species was dominated by C46, but also contained *Symbiodinium* closely related to B1 and B19, as well as B5.
